---
title: "Gene-set enrichment with regularized regression: simulation studies"
author: "Jitao David Zhang and Tao Fang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  BiocStyle::pdf_document:
    toc: true
vignette: >
    %\VignetteIndexEntry{simulation studies with gerr}
    %\VignetteEngine{knitr::rmarkdown}
---

# Background

In the manuscript *Gene-set Enrichment with Regularized Regression* (`gerr` in short), we propose using regularized regression to model the relationship between $Y$, a dichotomous dependent variable indicating membership of genes in a set of genes of interest (GOI hereafter), and $\Omega$, a matrix of dichotomous variables indicating membership of genes in gene-sets that are potentially overlapping or even identical with each other.

In this document, we perform simulation studies to demonstrate the sensitivity and specificity of the `gerr` method, using the software package of the same name that we published along with the manuscript and the elastic net implemented in the R software package `glmnet`. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      fig.width=5, fig.height=5)
library(gerr) 
library(glmnet)
library(msigdbr)
library(MASS)
set.seed(1887)
```  

```{r fromRibiosUtils, echo=FALSE, include=FALSE}
pairwiseDist <- function(list, fun=jaccardIndex) {
  len <- length(list)
  res <- matrix(0, len, len)
  colnames(res) <- rownames(res) <- names(list)
  vals <- sapply(seq(from = 1, to = len - 1), function(i) {
    sapply(seq(from = i + 1, to = len), function(j) {
      do.call(fun, list(list[[i]], list[[j]]))
    })
  })
  vv <- unlist(vals)
  res[lower.tri(res)] <- vv
  res <- t(res) + res
  diag(res) <- do.call(fun, list(list[[1]], list[[1]]))
  return(res)
}
overlapCoefficient <- function(x,y, checkUniqueNonNA=FALSE) {
  if(checkUniqueNonNA) {
    x <- uniqueNonNA(x)
    y <- uniqueNonNA(y)
  }
  res <- length(intersect(x,y))/pmin(length(x), length(y))
  return(res)
}
pairwiseOverlapCoefficient <- function(list) {
  return(pairwiseDist(list, fun=overlapCoefficient))
}
```

# Data used for simulation

We wish to use real-world gene-sets that are commonly used by the community for the simulation study for `gerr`, because synthesized gene-sets may have distributions of sizes, defined by the number of unique genes and overlapping patterns that depart from real-world gene-sets. Therefore we decided to use curated gene-sets provided by the MSigDB database (Category `C2`) for this purpose. These gene-sets are curated from online pathway databases, publications in PubMed, and collected knowledge of domain experts. Therefore they are not structured in tree structures like gene-sets from Gene Ontology (GO) or from the Reactome pathway database. 

```{r gmt}
simTibble <- msigdbr::msigdbr(species="Homo sapiens",
                             category = c("C2"))
simGenesAll <- with(simTibble, tapply(human_gene_symbol, gs_name, unique))
simGenesets <- sample(simGenesAll, 500)
simLen <- sapply(simGenesets, function(x) length(x))
```

The gene-sets are retrieved using the `msigdbr` package. For the purpose of simulation, we randomly sample 500 gene-sets from all gene-sets of the category C2 (N=`r length(simGenesets)`). The histogram below shows the distribution of gene-set size in the sub-sampled set of gene-sets.

```{r sizeDist}
{
  hist(simLen, xlab="Number of unique genes", 
       breaks=50, freq = FALSE,
       col="lightblue", main="Gene-set size distribution")
  lines(density(simLen, from=0), col="#004495", lwd=2)
}
```

Next, we investigate the degree of redundancy among these gene-sets. We use the overlap coefficient, defined by $|A \cap B|/min(|A|,|B|)$ between two sets $A$ and $B$, to measure this.

```{r overlapDist}
simPairwiseOverlap <- pairwiseOverlapCoefficient(simGenesets)
utOver <- simPairwiseOverlap[upper.tri(simPairwiseOverlap, diag=FALSE)]
{
  hist(utOver, xlab="Pairwise overlapping cofficient between gene-sets", 
       breaks=50, freq = FALSE,
       col="orange", 
       main="Overlapping coefficient distribution")
  lines(density(utOver, from=0), col="red", lwd=2)
}
```

```{r overlapStat, include=FALSE}
nonZeroOverlap <- mean(utOver!=0)
nonZeroUts <- utOver[utOver!=0]
nzutMed <- median(nonZeroUts)
nzutMad <- mad(nonZeroUts)
nzutPos <- mean(nonZeroUts>0.5)
```

We calculated the overlap coefficient for all unique pairs of gene-sets, and its distribution is shown in the histogram above. While no overlapping genes are identified between 69.8% pairs of gene-sets, overlapping genes are found in 30.2% pairs of gene-sets, with an median overlap coeffiecient of 0.047 and median absolute deviation (MAD) of 0.035. In about 0.5% of the cases, the overlap coefficient is equal or larger than 0.5. It suggests that while the assumption of independence between gene-sets may hold true when few gene-sets are used, there will be violation when many gene-sets are used for testing. 
# Simulations for specificity testing

We first start with the simplest simulation that is possible: we randomly select genes of one gene-set as the genes of interest (GOI), and test whether we can recover the gene-set using `gerr`.

First we convert the gene-sets into gene-pathway matrix required by `gerr::regression_selected_pathways`.

```{r gsMatrix}
bgGenes <- unique(unlist(simGenesets))
gsMatrix <- Matrix(sapply(simGenesets, function(x) as.integer(bgGenes %in% x)),
                   sparse=TRUE,
                   dimnames=list(bgGenes, names(simGenesets)))
```

```{r}
selInd <- sample(seq(along=simGenesets), 1)
selGsName <- names(simGenesets)[selInd]
selGs <- simGenesets[[selInd]]
selRes <- regression_selected_pathways(gene_input=selGs, gene_pathway_matrix = gsMatrix)
```

# Simulations for sensitivity testing

# R session info

```{r session}
sessionInfo()
```
