---
title: "Gene-set enrichment with regularized regression: simulation studies"
author: "Jitao David Zhang and Tao Fang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  BiocStyle::pdf_document:
    toc: true
vignette: >
    %\VignetteIndexEntry{simulation studies with gerr}
    %\VignetteEngine{knitr::rmarkdown}
---

# Background

In the manuscript *Gene-set Enrichment with Regularized Regression* (`gerr` in short), we propose using regularized regression to model the relationship between $Y$, a dichotomous dependent variable indicating membership of genes in a set of genes of interest (GOI hereafter), and $\Omega$, a matrix of dichotomous variables indicating membership of genes in gene-sets that are potentially overlapping or even identical with each other.

In this document, we perform simulation studies to demonstrate the sensitivity and specificity of the `gerr` method, using the software package of the same name that we published along with the manuscript and the elastic net implemented in the R software package `glmnet`. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE)
library(gerr) 
library(glmnet)
set.seed(1887)
```  

# Simulations for specificity testing

# Simulations for sensitivity testing

# R session info

```{r session}
sessionInfo()
```
